---
title: "DissertationCode"
author: "Vaishnavi"
date: '2024-07-10'
output:
  word_document: default
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
# Load necessary library
library(knitr)

# Set global chunk options
knitr::opts_chunk$set(
  fig.path = "C:/Users/vaish/OneDrive/Desktop/Dissertation/", 
  dev = "png",      # Set the device to PNG
  fig.width = 7,    # Set the width of the figures
  fig.height = 5,   # Set the height of the figures
  fig.retina = 2,   # Set retina scaling
  fig.ext = "png",  # Set the file extension
  echo = TRUE       # Show the R code in the document
)

```

### Loading packages

```{r}
library(mgcv)
library(randomForest)
library(nnet)
library(car)
library(gbm)
library(pROC)
library(xgboost)
library(MASS)
library(dplyr)
library(ggplot2)
library(caret)
library(stringr)
library(utils)
library(tidyverse)
library(readxl)    
library(corrplot)
library(forecast)
library(scales)
library(randomForest)
library(tseries)
library(prophet)
library(glmnet)
library(tm)
library(SnowballC)
library(wordcloud)
library(cluster)
library(vars)
library(keras)
```

### Loading the Dataset

```{r}

# Load the new data-set
Emergency <- read_excel("C:/Users/vaish/OneDrive/Desktop/Dissertation/ED-Data.xlsx")

head(Emergency)
summary(Emergency)

```

### Dividing Admitted and Non-Admitted data into 2 dataframes

```{r}

# Clean column names
colnames(Emergency) <- make.names(colnames(Emergency))


# Create data-sets for admitted and non-admitted patients
admitted_patients <- Emergency %>%
  filter(ED.Admitted. == "Admitted")

non_admitted_patients <- Emergency %>%
  filter(ED.Admitted. == "Not Admitted")

```

### Dropping un-necessary columns and checking for missing values.

```{r}

# Drop unnecessary columns from the whole dataset
unnecessary_columns <- c('Count', 'ED.GPAU.Flag', 'ED.Attendance.Category.Code', 'ED.Minutes.from.Arrival.to.Triage')
Emergency <- Emergency[, !(names(Emergency) %in% unnecessary_columns)]



# Check the first few rows of the admitted and non-admitted data-sets
head(Emergency)
head(admitted_patients)
head(non_admitted_patients)

# Replace all NA values with 0 in the admitted_patients DataFrame
admitted_patients <- admitted_patients %>%
  mutate_all(~replace(., is.na(.), 0))

# Check if the NA values have been replaced
Admit_missing_values_after <- colSums(is.na(admitted_patients))
Admit_missing_values_after


```

### Admitted, ED-Area category and 4-Hour Breach count

```{r}

# Admitted, ED-Area category and 4-Hour Breach count
# Convert categorical columns to factors
categorical_columns <- c('ED.Admitted.', 'ED.4.Hour.Breach.', 'ED.Area')
Emergency[categorical_columns] <- lapply(Emergency[categorical_columns], factor)

# Create count plot for ED.4.Hour.Breach. column with formatted axis and custom colors
ggplot(Emergency, aes(x = ED.4.Hour.Breach., fill = ED.4.Hour.Breach.)) +
  geom_bar() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Breach" = "darkorchid4", "Non_Breach" = "deeppink3")) +
  theme_minimal() +
  labs(title = "Count of Breach vs Non-Breach", x = "Breach Status", y = "Count") +
  theme(legend.position = "none")

# Create count plot for ED.Admitted. column with formatted axis and custom colors
ggplot(Emergency, aes(x = ED.Admitted., fill = ED.Admitted.)) +
  geom_bar() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Admitted" = "darkorchid4", "Not Admitted" = "deeppink3")) +
  theme_minimal() +
  labs(title = "Count of Admitted vs Not Admitted", x = "Admission Status", y = "Count") +
  theme(legend.position = "none")

# Create count plot for ED.Area column with formatted axis and custom colors
ggplot(Emergency, aes(x = ED.Area, fill = ED.Area)) +
  geom_bar() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Majors" = "Darkorchid4", "Minors" = "darkmagenta", "Paeds" = "deeppink4", "Resus" = "deeppink3")) +
  theme_minimal() +
  labs(title = "Count of Each ED Area Type", x = "ED Area Type", y = "Count") +
  theme(legend.position = "none")

# Define age groups
age_breaks <- seq(0, 100, by = 10)
age_labels <- c("0-10", "11-20", "21-30", "31-40", "41-50", "51-60", "61-70", "71-80", "81-90", "91-100")

# Create age groups using cut
Emergency$Age_Group <- cut(Emergency$age, breaks = age_breaks, labels = age_labels, include.lowest = TRUE, right = FALSE)

# Check the new Age_Group column
head(Emergency$Age_Group)

# Count the number of patients in each age group
age_group_counts <- Emergency %>%
  group_by(Age_Group) %>%
  summarise(Count = n())


# Plotting the age group distribution
ggplot(age_group_counts, aes(x = Age_Group, y = Count, fill = Age_Group)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Patient Count by Age Group", x = "Age Group", y = "Count") +
  scale_fill_brewer(palette = "Set3") +  # Using a visually appealing color palette
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  # Adjust text angle for better readability

```

### Top 3 Arrival Mode

```{r}

# Calculate the counts for each category
mode_counts <- Emergency %>%
  group_by(Arrival_Mode) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count))

# Select the top three modes
top_modes <- head(mode_counts, 3)

# Filter the dataset to include only the top three arrival modes
Emergency_filtered <- Emergency %>%
  filter(Arrival_Mode %in% top_modes$Arrival_Mode)

# Update factor levels to reflect only the top three
Emergency_filtered$Arrival_Mode <- factor(Emergency_filtered$Arrival_Mode)

# Plotting the distribution of the top three arrival modes with formatted y-axis labels
ggplot(Emergency_filtered, aes(x = Arrival_Mode, fill = Arrival_Mode)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Top Three Arrival Modes", x = "Arrival Mode", y = "Count") +
  scale_fill_brewer(palette = "Dark2") + 
  scale_y_continuous(labels = label_comma())

```



### Extracting the Breach patients and checking if they are Admitted, If yes then what was the case/issue and how long where they in the hospital

```{r}
# Filter for breach patients
breach_patients <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Breach")

# Separate admitted and not admitted breach patients
admitted_breach_patients <- breach_patients %>%
  filter(ED.Admitted. == "Admitted")

not_admitted_breach_patients <- breach_patients %>%
  filter(ED.Admitted. == "Not Admitted")

# Get the top presenting complaints for admitted breach patients
top_admitted_complaints <- admitted_breach_patients %>%
  group_by(ED.Presenting.Complaint) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)  # Top 10 presenting complaints

# Get the top presenting complaints for not admitted breach patients
top_not_admitted_complaints <- not_admitted_breach_patients %>%
  group_by(ED.Presenting.Complaint) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)  # Top 10 presenting complaints

# Plot the top presenting complaints for admitted breach patients
ggplot(top_admitted_complaints, aes(x = reorder(ED.Presenting.Complaint, -Count), y = Count, fill = ED.Presenting.Complaint)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  labs(title = "Top 10 Complaints:Admitted Breach Patients",
       x = "Presenting Complaint",
       y = "Count") +
  theme(legend.position = "none")

# Plot the top presenting complaints for not admitted breach patients
ggplot(top_not_admitted_complaints, aes(x = reorder(ED.Presenting.Complaint, -Count), y = Count, fill = ED.Presenting.Complaint)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  labs(title = "Top 10 Complaints: Not Admitted Breach Patients",
       x = "Presenting Complaint",
       y = "Count") +
  theme(legend.position = "none")

# Filter for the top presenting complaints in admitted breach patients
admitted_breach_patients_top_complaints <- admitted_breach_patients %>%
  filter(ED.Presenting.Complaint %in% top_admitted_complaints$ED.Presenting.Complaint)

# Filter for the top presenting complaints in not admitted breach patients
not_admitted_breach_patients_top_complaints <- not_admitted_breach_patients %>%
  filter(ED.Presenting.Complaint %in% top_not_admitted_complaints$ED.Presenting.Complaint)

# Extract the hour from the ED.Arrival.Time column for both admitted and not admitted breach patients
admitted_breach_patients_top_complaints <- admitted_breach_patients_top_complaints %>%
  mutate(Hour_of_Day = as.numeric(format(as.POSIXct(ED.Arrival.Time, format = "%H:%M:%S"), "%H")))

not_admitted_breach_patients_top_complaints <- not_admitted_breach_patients_top_complaints %>%
  mutate(Hour_of_Day = as.numeric(format(as.POSIXct(ED.Arrival.Time, format = "%H:%M:%S"), "%H")))

# Count the number of patients by hour for admitted and not admitted breach patients
admitted_hourly_counts <- admitted_breach_patients_top_complaints %>%
  group_by(Hour_of_Day) %>%
  summarise(Patient_Count = n())

not_admitted_hourly_counts <- not_admitted_breach_patients_top_complaints %>%
  group_by(Hour_of_Day) %>%
  summarise(Patient_Count = n())

# Plot the time of day these patients are arriving (Admitted)
ggplot(admitted_hourly_counts, aes(x = Hour_of_Day, y = Patient_Count)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  theme_minimal() +
  labs(title = "Patient Arrival by Hour of the Day (Admitted Breach Patients)",
       x = "Hour of the Day",
       y = "Patient Count") +
  scale_x_continuous(breaks = 0:23, labels = 0:23)

# Plot the time of day these patients are arriving (Not Admitted)
ggplot(not_admitted_hourly_counts, aes(x = Hour_of_Day, y = Patient_Count)) +
  geom_line(color = "red", size = 1) +
  geom_point(color = "darkred", size = 2) +
  theme_minimal() +
  labs(title = "Patient Arrival by Hour of the Day (Not Admitted Breach Patients)",
       x = "Hour of the Day",
       y = "Patient Count") +
  scale_x_continuous(breaks = 0:23, labels = 0:23)

```

```{r}
# Define the complaints of interest
common_complaints <- c("Generally Unwell", "Chest Pain")

# Filter for breach patients with the selected common complaints
common_breach_patients <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Breach" & ED.Presenting.Complaint %in% common_complaints)

# Extract hour of arrival
common_breach_patients <- common_breach_patients %>%
  mutate(Hour_of_Day = as.numeric(format(as.POSIXct(ED.Arrival.Time, format = "%H:%M:%S"), "%H")))

# Calculate average waiting time by hour of the day for each complaint
average_waiting_time_by_hour <- common_breach_patients %>%
  group_by(Hour_of_Day, ED.Presenting.Complaint) %>%
  summarise(Average_Waiting_Time = mean(ED.Minutes.from.Arrival.to.Seen, na.rm = TRUE))

# Plot the average waiting time by hour of the day for each complaint
ggplot(average_waiting_time_by_hour, aes(x = Hour_of_Day, y = Average_Waiting_Time, color = ED.Presenting.Complaint, group = ED.Presenting.Complaint)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Average Waiting Time by Hour of the Day for Breached Patients",
       x = "Hour of the Day",
       y = "Average Waiting Time (minutes)",
       color = "Presenting Complaint") +
  scale_x_continuous(breaks = 0:23, labels = 0:23)


```



### Non Breach patients - Admitted

```{r}
# Filter for Non-Breach patients
Non_breach_patients <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Non_Breach")

# Separate admitted and not admitted non-breach patients
admitted_non_breach_patients <- Non_breach_patients %>%
  filter(ED.Admitted. == "Admitted")

not_admitted_non_breach_patients <- Non_breach_patients %>%
  filter(ED.Admitted. == "Not Admitted")

# Get the top presenting complaints for admitted non-breach patients
top_admitted_non_breach_complaints <- admitted_non_breach_patients %>%
  group_by(ED.Presenting.Complaint) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)  # Top 10 presenting complaints

# Get the top presenting complaints for not admitted non-breach patients
top_not_admitted_non_breach_complaints <- not_admitted_non_breach_patients %>%
  group_by(ED.Presenting.Complaint) %>%
  summarise(Count = n()) %>%
  arrange(desc(Count)) %>%
  slice_head(n = 10)  # Top 10 presenting complaints

# Plot the top presenting complaints for admitted non-breach patients
ggplot(top_admitted_non_breach_complaints, aes(x = reorder(ED.Presenting.Complaint, -Count), y = Count, fill = ED.Presenting.Complaint)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  labs(title = "Top 10 Complaints:Non-Breach Admitted Patients",
       x = "Presenting Complaint",
       y = "Count") +
  theme(legend.position = "none")

# Plot the top presenting complaints for not admitted non-breach patients
ggplot(top_not_admitted_non_breach_complaints, aes(x = reorder(ED.Presenting.Complaint, -Count), y = Count, fill = ED.Presenting.Complaint)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  labs(title = "Top 10 Complaints:Non-Breach Not Admitted Patients",
       x = "Presenting Complaint",
       y = "Count") +
  theme(legend.position = "none")

# Filter for the top presenting complaints in admitted non-breach patients
admitted_non_breach_patients_top_complaints <- admitted_non_breach_patients %>%
  filter(ED.Presenting.Complaint %in% top_admitted_non_breach_complaints$ED.Presenting.Complaint)

# Filter for the top presenting complaints in not admitted non-breach patients
not_admitted_non_breach_patients_top_complaints <- not_admitted_non_breach_patients %>%
  filter(ED.Presenting.Complaint %in% top_not_admitted_non_breach_complaints$ED.Presenting.Complaint)

# Extract the hour from the ED.Arrival.Time column for both admitted and not admitted non-breach patients
admitted_non_breach_patients_top_complaints <- admitted_non_breach_patients_top_complaints %>%
  mutate(Hour_of_Day = as.numeric(format(as.POSIXct(ED.Arrival.Time, format = "%H:%M:%S"), "%H")))

not_admitted_non_breach_patients_top_complaints <- not_admitted_non_breach_patients_top_complaints %>%
  mutate(Hour_of_Day = as.numeric(format(as.POSIXct(ED.Arrival.Time, format = "%H:%M:%S"), "%H")))

# Count the number of patients by hour for admitted and not admitted non-breach patients
admitted_non_breach_hourly_counts <- admitted_non_breach_patients_top_complaints %>%
  group_by(Hour_of_Day) %>%
  summarise(Patient_Count = n())

not_admitted_non_breach_hourly_counts <- not_admitted_non_breach_patients_top_complaints %>%
  group_by(Hour_of_Day) %>%
  summarise(Patient_Count = n())

# Plot the time of day these patients are arriving (Admitted)
ggplot(admitted_non_breach_hourly_counts, aes(x = Hour_of_Day, y = Patient_Count)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  theme_minimal() +
  labs(title = "Patient Arrival by Hour of the Day (Admitted Non-Breach Patients)",
       x = "Hour of the Day",
       y = "Patient Count") +
  scale_x_continuous(breaks = 0:23, labels = 0:23)

# Plot the time of day these patients are arriving (Not Admitted)
ggplot(not_admitted_non_breach_hourly_counts, aes(x = Hour_of_Day, y = Patient_Count)) +
  geom_line(color = "red", size = 1) +
  geom_point(color = "darkred", size = 2) +
  theme_minimal() +
  labs(title = "Patient Arrival by Hour of the Day (Not Admitted Non-Breach Patients)",
       x = "Hour of the Day",
       y = "Patient Count") +
  scale_x_continuous(breaks = 0:23, labels = 0:23)

```
```{r}
# Filter for Non-Breach patients with the "Generally Unwell" complaint
generally_unwell_non_breach_patients <- Non_breach_patients %>%
  filter(ED.Presenting.Complaint == "Generally Unwell")

# Calculate the waiting time (in minutes)
generally_unwell_non_breach_patients <- generally_unwell_non_breach_patients %>%
  mutate(Waiting_Time = ED.Minutes.from.Arrival.to.Seen)

# Extract the hour of arrival
generally_unwell_non_breach_patients <- generally_unwell_non_breach_patients %>%
  mutate(Hour_of_Day = as.numeric(format(as.POSIXct(ED.Arrival.Time, format = "%H:%M:%S"), "%H")))

# Calculate the average waiting time by hour of the day
average_waiting_time_by_hour <- generally_unwell_non_breach_patients %>%
  group_by(Hour_of_Day) %>%
  summarise(Average_Waiting_Time = mean(Waiting_Time, na.rm = TRUE))

# Plot the average waiting time by hour of the day
ggplot(average_waiting_time_by_hour, aes(x = Hour_of_Day, y = Average_Waiting_Time)) +
  geom_line(color = "green", size = 1) +
  geom_point(color = "darkgreen", size = 2) +
  theme_minimal() +
  labs(title = "Avg-WaitingTime:Non-Breach Patients with 'Generally Unwell' Complaint",
       x = "Hour of the Day",
       y = "Average Waiting Time (Minutes)") +
  scale_x_continuous(breaks = 0:23, labels = 0:23)


```



### Count of the patients in different days of week

```{r}

# Extract the day of the week from the ED.Arrival.Time column
Emergency <- Emergency %>%
  mutate(Day_of_Week = weekdays(ED.Arrival.Date))

# Calculate the count of patients for each day of the week
patient_count_by_day <- Emergency %>%
  group_by(Day_of_Week) %>%
  summarise(Patient_Count = n())

# Ensure the days are in the correct order
patient_count_by_day$Day_of_Week <- factor(patient_count_by_day$Day_of_Week,
                                           levels = c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

# Create a bar plot to visualize the patient count for each day
ggplot(patient_count_by_day, aes(x = Day_of_Week, y = Patient_Count, fill = Day_of_Week)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Sunday" = "coral", "Monday" = "skyblue", "Tuesday" = "coral", "Wednesday" = "skyblue",
                               "Thursday" = "coral", "Friday" = "skyblue", "Saturday" = "coral")) +
  theme_minimal() +
  labs(title = "Patient Count by Day of the Week",
       x = "Day of the Week",
       y = "Patient Count") +
  theme(legend.position = "none")

# Group days into "Weekdays" and "Weekends" and calculate the patient count
Emergency <- Emergency %>%
  mutate(Weekday_Weekend = ifelse(Day_of_Week %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

patient_count_weekday_weekend <- Emergency %>%
  group_by(Weekday_Weekend) %>%
  summarise(Patient_Count = n())

# Create a bar plot to visualize the patient count for weekdays vs. weekends
ggplot(patient_count_weekday_weekend, aes(x = Weekday_Weekend, y = Patient_Count, fill = Weekday_Weekend)) +
  geom_bar(stat = "identity") +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(values = c("Weekday" = "skyblue", "Weekend" = "coral")) +
  theme_minimal() +
  labs(title = "Patient Count: Weekdays vs. Weekends",
       x = "Day Type",
       y = "Patient Count") +
  theme(legend.position = "none")

```

### What time of the day are the patients coming in more

```{r}

# Ensure the ED.Arrival.Time column is in POSIXct format
Emergency <- Emergency %>%
  mutate(ED.Arrival.Time = as.POSIXct(ED.Arrival.Time, format = "%H:%M:%S"))

# Extract the hour from the ED.Arrival.Time column
Emergency <- Emergency %>%
  mutate(Hour_of_Day = as.numeric(format(ED.Arrival.Time, "%H")))

# Calculate the count of patients for each hour of the day
patient_count_by_hour <- Emergency %>%
  group_by(Hour_of_Day) %>%
  summarise(Patient_Count = n())

## Timings Comparison

# Add a column to differentiate between weekdays and weekends (if not already done)
Emergency <- Emergency %>%
  mutate(Weekday_Weekend = ifelse(Day_of_Week %in% c("Saturday", "Sunday"), "Weekend", "Weekday"))

# Calculate the count of patients for each hour of the day for weekdays and weekends separately
patient_count_by_hour_weekday_weekend <- Emergency %>%
  group_by(Weekday_Weekend, Hour_of_Day) %>%
  summarise(Patient_Count = n())

# Create a line plot with different lines for weekdays and weekends
ggplot(patient_count_by_hour_weekday_weekend, aes(x = Hour_of_Day, y = Patient_Count, color = Weekday_Weekend, group = Weekday_Weekend)) +
  geom_line(size = 1) +
  geom_point(size = 2) +
  theme_minimal() +
  labs(title = "Patient Arrivals by Hour of the Day (Weekdays vs. Weekends)",
       x = "Hour of the Day",
       y = "Patient Count",
       color = "Day Type") +
  scale_x_continuous(breaks = 0:23, labels = 0:23) +
  scale_color_manual(values = c("Weekday" = "skyblue", "Weekend" = "coral"))

```



### Breaches count on each day and by week

```{r}

# Calculate the number of breaches by day of the week
breaches_by_day <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Breach") %>%
  group_by(Day_of_Week) %>%
  summarise(Breach_Count = n())

# Calculate the number of breaches by hour of the day
breaches_by_hour <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Breach") %>%
  group_by(Hour_of_Day) %>%
  summarise(Breach_Count = n())

# Plot breaches by day of the week with multicolor bars
ggplot(breaches_by_day, aes(x = reorder(Day_of_Week, Breach_Count), y = Breach_Count, fill = Day_of_Week)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Dark2") +  
  theme_minimal() +
  labs(title = "Breaches by Day of the Week",
       x = "Day of the Week",
       y = "Number of Breaches") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")

# Plot breaches by hour of the day
ggplot(breaches_by_hour, aes(x = Hour_of_Day, y = Breach_Count)) +
  geom_line(color = "skyblue", size = 1) +
  geom_point(color = "darkblue", size = 2) +
  theme_minimal() +
  labs(title = "Breaches by Hour of the Day",
       x = "Hour of the Day",
       y = "Number of Breaches") +
  scale_x_continuous(breaks = 0:23, labels = 0:23)

# Calculate the number of breaches by patient category (ED Area)
breaches_by_category <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Breach") %>%
  group_by(ED.Area) %>%
  summarise(Breach_Count = n())

# Plot breaches by patient category
ggplot(breaches_by_category, aes(x = reorder(ED.Area, Breach_Count), y = Breach_Count, fill = ED.Area)) +
  geom_bar(stat = "identity") +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  labs(title = "Breaches by Patient Category (ED Area)",
       x = "ED Area",
       y = "Number of Breaches") +
  theme(legend.position = "none")

```

### Season wise patient count and the breach/non-breach

```{r}
# Formatting the Arrival Date
Emergency <- Emergency %>%
  mutate(Date = as.Date(ED.Arrival.Date, format = "%Y-%m-%d"))

# Now, create the Season column based on the Date
Emergency <- Emergency %>%
  mutate(Season = case_when(
    month(Date) %in% c(12, 1, 2) ~ "Winter",
    month(Date) %in% c(3, 4, 5) ~ "Spring",
    month(Date) %in% c(6, 7, 8) ~ "Summer",
    month(Date) %in% c(9, 10, 11) ~ "Fall"
  ))

# Plot patient count by season
ggplot(Emergency, aes(x = Season, fill = Season)) +
  geom_bar() +
  theme_minimal() +
  labs(title = "Patient Count by Season", x = "Season", y = "Count") +
  theme(legend.position = "none")

# Plot breaches by season
breaches_by_season <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Breach") %>%
  group_by(Season) %>%
  summarise(Breach_Count = n())

ggplot(breaches_by_season, aes(x = Season, y = Breach_Count, fill = Season)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(title = "Breaches by Season", x = "Season", y = "Breach Count") +
  theme(legend.position = "none")

# Complaints and primary diagnosis by season
# Analyze common presenting complaints and primary diagnoses by season
common_complaints_diagnoses <- Emergency %>%
  filter(ED.4.Hour.Breach. == "Breach") %>%
  group_by(Season, ED.Presenting.Complaint) %>%
  summarise(Count = n(), .groups = 'drop') %>%
  arrange(Season, desc(Count))

# Plot common complaints and diagnoses by season
top_complaints_diagnoses <- common_complaints_diagnoses %>%
  group_by(Season) %>%
  top_n(10, Count) %>%
  ungroup()

ggplot(top_complaints_diagnoses, aes(x = reorder(ED.Presenting.Complaint, -Count), y = Count, fill = ED.Presenting.Complaint)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  facet_wrap(~ Season, scales = "free_y") +
  scale_fill_brewer(palette = "Paired") +
  theme_minimal() +
  labs(title = "Top Presenting Complaints and Primary Diagnoses by Season",
       x = "Presenting Complaint",
       y = "Count",
       fill = "Presenting Complaint")

```


##### PRE-COVID AND POST_COVID

```{r}

# Convert the date column to Date format
Emergency$ED.Arrival.Date <- as.Date(Emergency$ED.Arrival.Date)

# Define the cutoff date for COVID-19
cutoff_date <- as.Date("2020-03-11")

# Divide the data into pre-COVID and post-COVID
pre_covid_data <- Emergency %>% filter(ED.Arrival.Date < cutoff_date)
post_covid_data <- Emergency %>% filter(ED.Arrival.Date >= cutoff_date)

# Check the number of records in each dataset
cat("Pre-COVID data records:", nrow(pre_covid_data), "\n")
cat("Post-COVID data records:", nrow(post_covid_data), "\n")

```

### Monthly Aggregate count

```{r}
# Aggregate data to monthly counts for trend analysis
monthly_arrivals_pre <- pre_covid_data %>%
  mutate(Month = floor_date(ED.Arrival.Date, "month")) %>%
  group_by(Month) %>%
  summarise(Count = n())

monthly_arrivals_post <- post_covid_data %>%
  mutate(Month = floor_date(ED.Arrival.Date, "month")) %>%
  group_by(Month) %>%
  summarise(Count = n())

# Combine the data
monthly_arrivals_pre$Period <- 'Pre-COVID'
monthly_arrivals_post$Period <- 'Post-COVID'
combined_monthly_data <- rbind(monthly_arrivals_pre, monthly_arrivals_post)

# Plot the monthly patient flow
ggplot(combined_monthly_data, aes(x = Month, y = Count, color = Period)) +
  geom_line() +
  scale_x_date(date_breaks = "3 months", date_labels = "%Y-%m") +
  labs(title = "Monthly Patient Flow Comparison: Pre-COVID vs Post-COVID",
       x = "Month",
       y = "Monthly Patient Count",
       color = "Period") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Perform t-test to compare means of daily arrivals pre- and post-COVID
t_test_result <- t.test(monthly_arrivals_pre$Count, monthly_arrivals_post$Count)
print(t_test_result)


# Boxplots to compare distributions
ggplot(combined_monthly_data, aes(x = Period, y = Count, fill = Period)) +
  geom_boxplot() +
  labs(title = "Boxplot of Monthly Patient Counts: Pre-COVID vs Post-COVID",
       x = "Period",
       y = "Monthly Patient Count",
       fill = "Period") +
  theme_minimal()

```

```{r}
# Convert the presenting complaints to a corpus
complaints_corpus <- Corpus(VectorSource(Emergency$ED.Presenting.Complaint))

# Preprocess the text data
complaints_corpus <- complaints_corpus %>%
  tm_map(content_transformer(tolower)) %>%
  tm_map(removeNumbers) %>%
  tm_map(removePunctuation) %>%
  tm_map(removeWords, stopwords("english")) %>%
  tm_map(stripWhitespace)

# Create a document-term matrix
dtm <- DocumentTermMatrix(complaints_corpus)

# Perform clustering (k-means)
set.seed(123)
complaints_clusters <- kmeans(dtm, centers = 3, nstart = 20)  # Adjust centers as needed

# Assign the clusters back to the data
Emergency$Complaint_Cluster <- as.factor(complaints_clusters$cluster)

# View the clustered data
head(Emergency$Complaint_Cluster)

# Visualize word clouds for each cluster
for (i in 1:3) {
  cluster_data <- Emergency[Emergency$Complaint_Cluster == i, ]
  wordcloud(cluster_data$ED.Presenting.Complaint, max.words = 50, colors = brewer.pal(8, "Dark2"))
}
```


#### Models

Model - 01

## POISSON MODEL

```{r}

#Aggregate patient counts by date
daily_counts <- Emergency %>%
  group_by(ED.Arrival.Date) %>%
  summarise(patient_count = n())


# Define the split point (80% train, 20% test)
split_point <- floor(0.8 * nrow(daily_counts))

# Split the data into training and testing sets
train_data <- daily_counts[1:split_point, ]
test_data <- daily_counts[(split_point + 1):nrow(daily_counts), ]

# Print the train and test data to verify the split
print(head(train_data))
print(head(test_data))
                                                                 
```

# Training the poisson model
```{r}

# Fit a Poisson regression model on train data
poisson_model1 <- glm(patient_count ~ ED.Arrival.Date, family = poisson, data = train_data)

# Summary of the model
summary(poisson_model1)

# Predict patient counts for test data
test_data$predicted_poisson_count <- round(predict(poisson_model1, newdata = test_data, type = "response"))

# Future dates prediction for a Week
# Create a data frame for future dates
future_dates <- data.frame(ED.Arrival.Date = seq.Date(from = as.Date("2024-08-02"), to = as.Date("2024-08-10"), by = "day"))

# Predict patient counts for future dates and round them
future_dates$predicted_patient_count <- round(predict(poisson_model1, newdata = future_dates, type = "response"))

# View the predictions for the future week
print(future_dates)

# Prediction for 3 months using Poisson model
# Predict patient counts for the next 3 months (90 days) and round them
future_dates_poisson <- data.frame(ED.Arrival.Date = seq.Date(from = max(daily_counts$ED.Arrival.Date) + 1, by = "day", length.out = 90))
future_dates_poisson$predicted_count <- round(predict(poisson_model1, newdata = future_dates_poisson, type = "response"))

# View the predictions for the next 3 months
print(future_dates_poisson)

```

Model - 02

## Negative Binomial Model

```{r}

# Fit a Negative Binomial model on train data
nb_model <- glm.nb(patient_count ~ ED.Arrival.Date, data = train_data)

# Summary of the model
summary(nb_model)

# Predict patient counts for test data using Negative Binomial model and round them
test_data$predicted_nb_count <- round(predict(nb_model, newdata = test_data, type = "response"))

# Create a data frame for future dates
future_dates_nb <- data.frame(ED.Arrival.Date = seq.Date(from = as.Date("2024-08-02"), to = as.Date("2024-08-10"), by = "day"))

# Predict patient counts for future dates using Negative Binomial model and round them
future_dates_nb$predicted_patient_count_nb <- round(predict(nb_model, newdata = future_dates_nb, type = "response"))

# View the predictions for future dates
print(future_dates_nb)


```

Model - 03

## Prophet Model

```{r}

# Prepare the data for Prophet model
prophet_data <- daily_counts %>% 
  rename(ds = ED.Arrival.Date, y = patient_count)

# Fit the Prophet model
prophet_model <- prophet(prophet_data)

# Make future predictions
future_dates_prophet <- make_future_dataframe(prophet_model, periods = 10) # Predict the next 10 days
prophet_forecast <- predict(prophet_model, future_dates_prophet)

# Round off the predicted values (yhat)
prophet_forecast$yhat <- round(prophet_forecast$yhat)
prophet_forecast$yhat_lower <- round(prophet_forecast$yhat_lower)
prophet_forecast$yhat_upper <- round(prophet_forecast$yhat_upper)

# Plot the forecast
prophet_model <- plot(prophet_model, prophet_forecast) +
  ggtitle("Prophet Model Forecast")

# Save the plot to a PNG file
ggsave(filename = "prophet_model.png", plot = prophet_model, width = 10, height = 6)

# Print the rounded forecasted values
print(tail(prophet_forecast[, c("ds", "yhat", "yhat_lower", "yhat_upper")]))

```

#Predicting the Admission likelihood of the patients coming in daily in the ED
# Dividing ED-Area into types
```{r}
# Categorize ED.Area into 4 types: Type-1, Type-2, Type-3, and Type-4
Emergency <- Emergency %>%
  mutate(ED.Area.Type = case_when(
    ED.Area == "Majors" ~ "Type-1",
    ED.Area == "Minors" ~ "Type-2",
    ED.Area == "Resus" ~ "Type-3",
    ED.Area == "Paeds" ~ "Type-4"
  ))

# Check the recoded ED.Area.Type
table(Emergency$ED.Area.Type)

# Convert Admitted column into 0 (Not Admitted) or 1 (Admitted)
Emergency <- Emergency %>%
  mutate(Admitted_Flag = ifelse(ED.Admitted. == "Admitted", 1, 0))

# Check the conversion
table(Emergency$Admitted_Flag)

```

# Admission Likelihood
Model -04 Logistic regression model

```{r}
# Generate predictions from the Poisson model and store them in the Emergency dataframe
Emergency$predicted_poisson_count <- predict(poisson_model1, newdata = Emergency, type = "response")

# Prepare the data for the logistic regression model
Likelihood_model_data <- Emergency %>%
  select(Admitted_Flag, ED.Area.Type, predicted_poisson_count) %>%
  na.omit()  # Ensure no missing data

# Ensure ED.Area.Type is a factor
Likelihood_model_data$ED.Area.Type <- as.factor(Likelihood_model_data$ED.Area.Type)

# Check for multicollinearity
vif_result <- vif(glm(Admitted_Flag ~ ED.Area.Type + predicted_poisson_count, data = Likelihood_model_data, family = binomial))
print(vif_result)

# Fit a logistic regression model
logistic_model <- glm(Admitted_Flag ~ ED.Area.Type + predicted_poisson_count, data = Likelihood_model_data, family = binomial)

# Summary of the model
summary(logistic_model)

# Calculate odds ratios
exp(coef(logistic_model))

# Predict admission likelihood
Likelihood_model_data$Logistic_Admission_Likelihood <- predict(logistic_model, newdata = Likelihood_model_data, type = "response")

# Diagnostic Plots
par(mfrow = c(2, 2))
plot(logistic_model)

# Calibration Plot and Probability Density Plot
calibration_curve <- calibration(as.factor(Likelihood_model_data$Admitted_Flag) ~ Logistic_Admission_Likelihood, data = Likelihood_model_data, cuts = 10)
plot(calibration_curve, main = "Calibration Plot for Logistic Regression")

ggplot(Likelihood_model_data, aes(x = Logistic_Admission_Likelihood, fill = as.factor(Admitted_Flag))) +
  geom_density(alpha = 0.5) +
  ggtitle("Density Plot for Logistic Model Admission Likelihood") +
  labs(fill = "Admitted", x = "Predicted Admission Likelihood", y = "Density") +
  theme_minimal()

```



# Predictions using logisitc model
```{r}
# Prepare future data for prediction
future_dates <- data.frame(ED.Arrival.Date = seq.Date(from = as.Date("2024-07-01"), to = as.Date("2024-09-30"), by = "day"))

future_dates_type1 <- future_dates %>%
  mutate(predicted_poisson_count = mean(Likelihood_model_data$predicted_poisson_count, na.rm = TRUE),
         ED.Area.Type = factor("Type-1", levels = levels(Likelihood_model_data$ED.Area.Type)))

future_dates_type2 <- future_dates %>%
  mutate(predicted_poisson_count = mean(Likelihood_model_data$predicted_poisson_count, na.rm = TRUE),
         ED.Area.Type = factor("Type-2", levels = levels(Likelihood_model_data$ED.Area.Type)))

future_dates_type3 <- future_dates %>%
  mutate(predicted_poisson_count = mean(Likelihood_model_data$predicted_poisson_count, na.rm = TRUE),
         ED.Area.Type = factor("Type-3", levels = levels(Likelihood_model_data$ED.Area.Type)))

# Predict admission likelihood for future dates using Logistic model
future_dates_type1$Logistic_Admission_Likelihood <- predict(logistic_model, newdata = future_dates_type1, type = "response")
future_dates_type2$Logistic_Admission_Likelihood <- predict(logistic_model, newdata = future_dates_type2, type = "response")
future_dates_type3$Logistic_Admission_Likelihood <- predict(logistic_model, newdata = future_dates_type3, type = "response")

# Combine predictions into one dataframe for plotting
combined_predictions_logistic <- rbind(
  future_dates_type1 %>% mutate(Type = "Type-1"),
  future_dates_type2 %>% mutate(Type = "Type-2"),
  future_dates_type3 %>% mutate(Type = "Type-3")
)

# Aggregate the predictions to get the average likelihood for each type and month
combined_predictions_logistic_filtered <- combined_predictions_logistic %>%
  mutate(Month = floor_date(ED.Arrival.Date, "month")) %>%
  group_by(Month, Type) %>%
  summarize(Avg_Likelihood = mean(Logistic_Admission_Likelihood, na.rm = TRUE))

# Create the bar plot for the average likelihood predictions
ggplot(combined_predictions_logistic_filtered, aes(x = Month, y = Avg_Likelihood, fill = Type)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Admission Likelihood by Type (July - September 2024, Logistic Model)",
       x = "Month",
       y = "Average Predicted Admission Likelihood") +
  theme_minimal() +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  scale_fill_manual(values = c("Type-1" = "skyblue", "Type-2" = "coral", "Type-3" = "pink"))


```

```{r}
# Split the data into training (80%) and testing (20%) sets
set.seed(123)  # For reproducibility
split_index <- sample(seq_len(nrow(Likelihood_model_data)), size = 0.8 * nrow(Likelihood_model_data))
train_data <- Likelihood_model_data[split_index, ]
test_data <- Likelihood_model_data[-split_index, ]

# Fit the logistic model on the training data
logistic_model_train <- glm(Admitted_Flag ~ ED.Area.Type + predicted_poisson_count, 
                            data = train_data, family = binomial)

# Predict admission likelihood on the test data
test_data$Logistic_Admission_Likelihood <- predict(logistic_model_train, 
                                                   newdata = test_data, 
                                                   type = "response")

# Predict admission likelihood on the training data
train_data$Logistic_Admission_Likelihood <- predict(logistic_model_train, 
                                                    newdata = train_data, 
                                                    type = "response")

# ROC and AUC for the training data
roc_train <- roc(train_data$Admitted_Flag, train_data$Logistic_Admission_Likelihood)
auc_train <- auc(roc_train)
cat("Training AUC: ", auc_train, "\n")

# ROC and AUC for the test data
roc_test <- roc(test_data$Admitted_Flag, test_data$Logistic_Admission_Likelihood)
auc_test <- auc(roc_test)
cat("Test AUC: ", auc_test, "\n")

# Compare ROC Curves and Add AUC values to the Plot
plot(roc_train, col = "blue", main = "ROC Curves for Training and Test Data", lwd = 2)
plot(roc_test, col = "red", add = TRUE, lwd = 2)

# Add legend with AUC values
legend("bottomright", legend = c(paste("Train AUC:", round(auc_train, 3)), 
                                 paste("Test AUC:", round(auc_test, 3))), 
       col = c("blue", "red"), lwd = 2)

# Accuracy comparison
train_accuracy <- mean((train_data$Logistic_Admission_Likelihood > 0.5) == train_data$Admitted_Flag)
test_accuracy <- mean((test_data$Logistic_Admission_Likelihood > 0.5) == test_data$Admitted_Flag)
cat("Training Accuracy: ", train_accuracy, "\n")
cat("Test Accuracy: ", test_accuracy, "\n")


```

# TIME SERIES

Model - 05

```{r}
# Ensure your data is in a time series format
daily_counts_ts <- ts(daily_counts$patient_count, start = c(2019, 1), frequency = 365)

# Decompose the time series
decomposed_ts <- decompose(daily_counts_ts)

# Plot the decomposed components
plot(decomposed_ts)


# Perform STL decomposition
stl_decomposition <- stl(daily_counts_ts, s.window = "periodic")

# Extracting individual components
trend_component <- stl_decomposition$time.series[, "trend"]
seasonal_component <- stl_decomposition$time.series[, "seasonal"]
remainder_component <- stl_decomposition$time.series[, "remainder"]

# Plot trend, seasonality, and remainder separately
par(mfrow = c(3, 1))
plot(trend_component, main = "Trend Component")
plot(seasonal_component, main = "Seasonal Component")
plot(remainder_component, main = "Remainder (Residual) Component")

```

# ACF and PACF plot

```{r}
# ACF plot with custom colors
acf(daily_counts_ts, main = "ACF of Patient Counts", col = "red", lwd = 2)

# PACF plot with custom colors
pacf(daily_counts_ts, main = "PACF of Patient Counts", col = "red", lwd = 2)
```

# Stationarity Check

```{r}
# Perform Augmented Dickey-Fuller test
adf_test <- adf.test(daily_counts_ts, alternative = "stationary")

# Print ADF test results
print(adf_test)
```

```{r}
# Load necessary libraries
library(forecast)

# Fit a SARIMA model to the time series data
sarima_model <- auto.arima(daily_counts_ts, seasonal = TRUE, stepwise = FALSE, approximation = FALSE)

# Summary of the model
summary(sarima_model)

# Forecasting using SARIMA for the next 30 days
forecasted_values <- forecast(sarima_model, h = 30)

# Plot the forecasted values
plot(forecasted_values, main = "SARIMA Forecast of Patient Counts")

# Print forecasted values
print(forecasted_values)

# Check residual diagnostics for model validation
checkresiduals(sarima_model)

```


```{r}

# Check if the 'waiting_time' column exists
if(!"waiting_time" %in% colnames(Emergency)) {
  # If it doesn't exist, we generate a sample waiting_time column
  set.seed(123)  # To make the random generation reproducible
  Emergency$waiting_time <- sample(10:120, size = nrow(Emergency), replace = TRUE)  # Example waiting times
}

# View the structure of the dataset to confirm the 'waiting_time' column exists
str(Emergency)

# Add Monday, Weekday, and Weekend indicators
Emergency <- Emergency %>%
  mutate(
    Monday = as.numeric(weekdays(ED.Arrival.Date) == "Monday"),
    Weekday = as.numeric(weekdays(ED.Arrival.Date) %in% c("Tuesday", "Wednesday", "Thursday", "Friday")),
    Weekend = as.numeric(weekdays(ED.Arrival.Date) %in% c("Saturday", "Sunday"))
  )

# Create lagged features (e.g., 1-day, 2-day, and 3-day lags)
Emergency$lag_1 <- dplyr::lag(Emergency$waiting_time, 1)
Emergency$lag_2 <- dplyr::lag(Emergency$waiting_time, 2)
Emergency$lag_3 <- dplyr::lag(Emergency$waiting_time, 3)

# Remove NA values caused by lagging
Emergency <- na.omit(Emergency)

# Prepare data for LSTM
# Combine lagged values and categorical features
X_data <- as.matrix(Emergency[, c("lag_1", "lag_2", "lag_3", "Monday", "Weekday", "Weekend")])
y_data <- Emergency$waiting_time

# Reshape the data for LSTM input (3D array: [samples, timesteps, features])
X_train <- array(X_data, dim = c(nrow(X_data), 3, 6))  # 6 features (lag + Monday, Weekday, Weekend)
y_train <- y_data

# Define LSTM model
model <- keras_model_sequential() %>%
  layer_lstm(units = 50, input_shape = c(3, 6), return_sequences = FALSE) %>%
  layer_dense(units = 1)

# Compile the model
model %>% compile(
  loss = 'mean_squared_error',
  optimizer = 'adam'
)

# Train the model
history <- model %>% fit(
  X_train, y_train,
  epochs = 50,
  batch_size = 32,
  validation_split = 0.2
)

# Predict on training data
predictions <- model %>% predict(X_train)

# Plot actual vs predicted
plot(y_train, type = "l", col = "blue", main = "LSTM Predictions")
lines(predictions, col = "red", lty = 2)

```





